
## Define library path where R looks for installed packages and load them
```{r, warning=F, message=F}
.libPaths("P://Pro00110219 - Predictive modeling using EHR/ch513/Rtools")
#install.packages ("dplyr", repo="http://archive.linux.duke.edu/cran/")
#install.packages ("torch", repo="http://archive.linux.duke.edu/cran/")
packageVersion("rlang")
library(RODBC)
library(dplyr)
library(lubridate)
library(ggplot2)
library(tidyr)
library(data.table)
library(tableone)
library(Hmisc)
library(caret)
library(survival)
library(torch)
```


## Stingent Cohort Logic:


## Phase 1:
## - Patients with outpatient visit between 2016 to 2018
## - Patients without CVD events before 2016

```{r, warning=FALSE}
## 1138417 patients
setwd("../../ch513/PREVENT/data")
load('pat_enroll.RData')
length(pat_enroll)
```


## Phase 2:
## - Patients with both SCRE and SBP

```{r}
## 605554 patients
setwd("../../ch513/PREVENT/data")
load('pat_list_lab.RData')
load('pat_list_vital.RData')
load('dat_lab_cvd1.RData')
dat_phase2 <- pat_enroll[pat_enroll %in% dat_lab_cvd1[dat_lab_cvd1$lab=="scre",]$PATID & pat_enroll %in% pat_list_sbp]
length(dat_phase2)
```



## Phase 3:
## - Both SCRE and SBP measured after 2016
## - Index Date: Patients have SCRE and SBP in same year, set Index Date as later date between SCRE and SBP
## - Patients with all 3 labs in 3 year lookback window: TC, HDL-C, BMI
## - Patients without CVD before index date


```{r}
## 518469 patients have both SCRE and SBP after 2016
## 509378 patients have SCRE and SBP in same year
setwd("../../ch513/PREVENT/data")
load('dat_sbp.RData')
dat_scre=dat_lab_cvd1[dat_lab_cvd1$lab=="scre",c("PATID", "SPECIMEN_DATE",'RESULT_NUM', "lab")]
colnames(dat_scre)[2:4]=c("DATE", "Result", "VAR")
dat_sbp_working=dat_sbp[,c("PATID", "MEASURE_DATE", 'vital_value_ave')]
dat_sbp_working=data.frame(dat_sbp_working, var="sbp")
colnames(dat_sbp_working)=colnames(dat_scre)

dat_scre_working=dat_scre %>%
  filter(as.Date(DATE) > as.Date('2015-12-31'))

dat_sbp_working=dat_sbp_working %>%
  filter(as.Date(DATE) > as.Date('2015-12-31'))

pat_both <- Reduce(intersect, list(unique(dat_sbp_working$PATID), 
                                   unique(dat_scre_working$PATID), 
                                   unique(dat_phase2)))

dat_scre_working=dat_scre_working %>%
  filter(PATID %in%pat_both)

dat_sbp_working=dat_sbp_working %>%
  filter(PATID %in%pat_both)

df_sbp=dat_sbp_working
df_scre=dat_scre_working

setDT(df_sbp)
setDT(df_scre)

df_sbp <- df_sbp[,.(patid=PATID, sbp_date = DATE, sbp_result = Result)]
df_scre <- df_scre[,.(patid=PATID, scre_date = DATE, scre_result = Result)]

df_sbp[, sbp_date := as.IDate(sbp_date)] 
df_scre[, scre_date := as.IDate(scre_date)] 

df_sbp <- unique(df_sbp)
df_scre <- unique(df_scre)

df_sbp[, `:=`(start_date = sbp_date - 365, end_date = sbp_date + 365)]

num_chunks <- ceiling(length(pat_both) / 1000)


dat_index <- NULL

for (i in 1:num_chunks) {
  print(i)
  ind <- 1:1000 + ((i - 1) * 1000)
  pat_tmp <- pat_both[ind]
  
  # filter data for the current chunk
  df_sbp_tmp <- df_sbp %>% filter(patid %in% pat_tmp)
  df_scre_tmp <- df_scre %>% filter(patid %in% pat_tmp)
  
  # inner join with the date range condition
  tmp <- inner_join(df_scre_tmp, df_sbp_tmp,
                    join_by(patid == patid, scre_date >= start_date, scre_date <= end_date))
  
  # get min date and absolute date difference
  tmp$col.min <- pmin(tmp$scre_date, tmp$sbp_date, na.rm = TRUE)
  tmp$col.diff <- abs(as.numeric(tmp$scre_date) - as.numeric(tmp$sbp_date))
  
  # filter rows with the minimum col.min for each PATID
  min_rows <- tmp[, .I[col.min == min(col.min, na.rm = TRUE)], by = patid]$V1
  tmp_sel <- tmp[min_rows]
  
  # filter rows with the minimum col.diff for each PATID
  min_rows <- tmp_sel[, .I[col.diff == min(col.diff, na.rm = TRUE)], by = patid]$V1
  tmp_sel <- tmp_sel[min_rows]
  
  # assign INDEX_DATE as the later of scre_date and sbp_date
  tmp_sel$INDEX_DATE <- pmax(tmp_sel$scre_date, tmp_sel$sbp_date, na.rm = TRUE)
  
  # subset to PATID and INDEX_DATE columns
  tmp_result <- tmp_sel[, .(PATID = patid, INDEX_DATE, scre = scre_result, sbp = sbp_result)]
  
  # append to the final result
  dat_index <- rbind(dat_index, tmp_result)
}
dat_index <- dat_index[, .SD[1], by = PATID]
length(dat_index$PATID)
```

```{r}
## 235005 patients have all the 3 labs from TC, HDLC, and BMI in 3 years lookback window from index date
setwd("../../ch513/PREVENT/data")
load('dat_bmi.RData')
load('dat_lab_cvd1.RData')

dat_index1 = dat_index[, c("PATID", "INDEX_DATE")]

# join from previous step: index date
dat_lab_cvd2 <- inner_join(
  dat_lab_cvd1[, c("PATID", "SPECIMEN_DATE", "lab", "LAB_LOINC", "RESULT_NUM", "RESULT_UNIT")], 
  dat_index1, 
  by = "PATID"
)

# pat with hdlc in 3 years lookback window: 236695 patients
dat_lab_hdlc=dat_lab_cvd2[which(dat_lab_cvd2$lab=="hdlc"),]
dat_lab_hdlc$diff=as.Date(dat_lab_hdlc$INDEX_DATE)-as.Date(dat_lab_hdlc$SPECIMEN_DATE)
dat_lab_hdlc$diff_year=dat_lab_hdlc$diff/365
dat_lab_hdlc=dat_lab_hdlc[which(is.na(dat_lab_hdlc$diff_year)!=1),]
dat_lab_hdlc=dat_lab_hdlc[which(dat_lab_hdlc$diff_year<3 & dat_lab_hdlc$diff_year>=0),]
pat_hdlc_lookback=unique(dat_lab_hdlc$PATID)

# pat with bmi in 3 years lookback window: 496539 patients
dat_vital_bmi=inner_join(data.frame(dat_bmi), dat_index1, by="PATID")
dat_vital_bmi=setDT(dat_vital_bmi)
dat_vital_bmi[, INDEX_DATE := as.Date(INDEX_DATE)]
dat_vital_bmi[, MEASURE_DATE := as.Date(MEASURE_DATE)]
dat_vital_bmi[, diff_year := as.numeric(difftime(INDEX_DATE, MEASURE_DATE, units = "days"))/365]
dat_vital_bmi=dat_vital_bmi[which(is.na(dat_vital_bmi$diff_year)!=1),]
dat_vital_bmi <- dat_vital_bmi[!is.na(diff_year)]
dat_vital_bmi <- dat_vital_bmi[diff_year < 3]
dat_vital_bmi <- dat_vital_bmi[diff_year >=0]
pat_bmi_lookback=unique(dat_vital_bmi$PATID)

# pat with tc in 3 years lookback window: 237306 patients
dat_tc=dat_lab_cvd2[which(dat_lab_cvd2$lab=="tc"),]
dat_tc=setDT(dat_tc)
dat_tc[, INDEX_DATE := as.Date(INDEX_DATE)]
dat_tc[, SPECIMEN_DATE := as.Date(SPECIMEN_DATE)]
dat_tc[, diff_year := as.numeric(difftime(INDEX_DATE, SPECIMEN_DATE, units = "days"))/365]
dat_tc=dat_tc[which(is.na(dat_tc$diff_year)!=1),]
dat_tc <- dat_tc[!is.na(diff_year)]
dat_tc <- dat_tc[diff_year < 3]
dat_tc <- dat_tc[diff_year >=0]
pat_tc_lookback=unique(dat_tc$PATID)

# overall lookback: 235005 patients
pat_all_lookback <- Reduce(intersect, list(pat_hdlc_lookback, pat_bmi_lookback, pat_tc_lookback))

# get patients in all three lookback groups
dat_lab_lookback <- dat_lab_cvd2[dat_lab_cvd2$PATID %in% pat_all_lookback, ]

length(pat_all_lookback)
```


```{r}
## 229798 patients don't have cvd before index date
# no cvd before index date
load('../../ch513/PREVENT/data/dat_diagnosis_cvd1.RData')
dat_lab_lookback_unique <- dat_lab_lookback %>% distinct(PATID, INDEX_DATE)
dat_cvd_lookback = inner_join(data.frame(dat_diagnosis_cvd1), dat_lab_lookback_unique, by="PATID")
dat_cvd_lookback$diff=as.Date(dat_cvd_lookback$INDEX_DATE)-as.Date(dat_cvd_lookback$ADMIT_DATE)
dat_cvd_lookback$diff_year=dat_cvd_lookback$diff/365
dat_cvd_lookback=dat_cvd_lookback[which(is.na(dat_cvd_lookback$diff_year)!=1),]
dat_cvd_lookback=dat_cvd_lookback[which(dat_cvd_lookback$diff_year>=0),]
pat_cvd_lookback=unique(dat_cvd_lookback$PATID)

# exclude PATIDs in pat_cvd_lookback
dat_phase3 <- dat_lab_lookback[!(dat_lab_lookback$PATID %in% pat_cvd_lookback), ]
length(unique(dat_phase3$PATID))
```



## Phase 4

## - Patients with age, sex and race
## - Patients have at least one followup at 7 days after the index date 
## - Patients have index date within 2016-2018


```{r}
## 229798 patients have both age, sex and race info
load('../../ch513/PREVENT/data/dat_demo.RData')
pat_age <- dat_demo[!is.na(BIRTH_DATE), .N, by = .(PATID)]$PATID
pat_sex <- dat_demo[!is.na(SEX), .N, by = .(PATID)]$PATID
pat_race <- dat_demo[!is.na(RACE), .N, by = .(PATID)]$PATID
pat_demo <- unique(Reduce(intersect, list(pat_age, pat_sex, pat_race)))

# exclude PATIDs without demographic data
dat_phase4_temp <- dat_phase3[dat_phase3$PATID %in% pat_demo, ]
length(unique(dat_phase4_temp$PATID))
```

```{r}
# 220915 patients have at least 1 follow up at 7 days after index date
load('../../ch513/PREVENT/data/dat_followup_wide.RData')

# get unique PATID and INDEX_DATE from dat_demo
dat_followup <- dat_phase4_temp %>%
  distinct(PATID, INDEX_DATE)

# join with dat_followup_wide
dat_phase4_joined <- inner_join(dat_followup, dat_followup_wide, by = "PATID")

# at least 1 follow up at 7 days after index date
dat_phase4_joined$followup_days_since_index = as.Date(dat_phase4_joined$end_date)-as.Date(dat_phase4_joined$INDEX_DATE)
dat_phase4_joined = dat_phase4_joined[dat_phase4_joined$followup_days_since_index>=7, ]
dat_phase4_follow = dat_phase4_temp[dat_phase4_temp$PATID %in% dat_phase4_joined$PATID, ]
length(unique(dat_phase4_follow$PATID))
```

```{r}
# 181337 patients have index date within 2016-2018
dat_phase4 = dat_phase4_follow[dat_phase4_follow$INDEX_DATE<"2019-01-01" & dat_phase4_follow$INDEX_DATE>="2016-01-01", ]
length(unique(dat_phase4$PATID))
```


## Phase 5:

## clean data for each patient in final strict cohort

## Labs

```{r}
cohort = dat_index[dat_index$PATID %in% dat_phase4$PATID,]
  
# tc
dat_lab_tc=dat_phase4[which(dat_phase4$lab=="tc"),]
setDT(dat_lab_tc)
dat_lab_tc[, diff := as.Date(INDEX_DATE) - as.Date(SPECIMEN_DATE)]
dat_lab_tc[, diff_year := diff / 365]
dat_lab_tc <- dat_lab_tc[!is.na(diff_year) & diff_year < 3 & diff_year >= 0]
dat_lab_tc_min_diff <- dat_lab_tc[order(PATID, diff)][, .SD[1], by = PATID]
dat_lab_tc_min_diff <- dat_lab_tc_min_diff[, .(PATID, tc = RESULT_NUM)]


# hdlc
dat_lab_hdlc=dat_phase4[which(dat_phase4$lab=="hdlc"),]
setDT(dat_lab_hdlc)
dat_lab_hdlc[, diff := as.Date(INDEX_DATE) - as.Date(SPECIMEN_DATE)]
dat_lab_hdlc[, diff_year := diff / 365]
dat_lab_hdlc <- dat_lab_hdlc[!is.na(diff_year) & diff_year < 3 & diff_year >= 0]
dat_lab_hdlc_min_diff <- dat_lab_hdlc[order(PATID, diff)][, .SD[1], by = PATID]
dat_lab_hdlc_min_diff <- dat_lab_hdlc_min_diff[, .(PATID, hdlc = RESULT_NUM)]

# bmi
dat_vital_bmi=inner_join(data.frame(dat_bmi), dat_index1, by="PATID")
dat_vital_bmi = dat_vital_bmi[dat_vital_bmi$PATID %in% dat_phase4$PATID,]
dat_vital_bmi=setDT(dat_vital_bmi)
dat_vital_bmi[, diff := as.Date(INDEX_DATE) - as.Date(MEASURE_DATE)]
dat_vital_bmi[, diff_year := diff / 365]
dat_vital_bmi <- dat_vital_bmi[!is.na(diff_year) & diff_year < 3 & diff_year >= 0]
dat_vital_bmi_min_diff <- dat_vital_bmi[order(PATID, diff)][, .SD[1], by = PATID]
dat_vital_bmi_min_diff <- dat_vital_bmi_min_diff[, .(PATID, bmi = vital_value_ave)]


# join all
cohort <- left_join(cohort, dat_lab_tc_min_diff, by = "PATID")
cohort <- left_join(cohort, dat_lab_hdlc_min_diff, by = "PATID")
cohort <- left_join(cohort, dat_vital_bmi_min_diff, by = "PATID")

# check: no one row have all 3 labs missing at the same time
cohort %>%
  filter(is.na(tc) | is.na(hdlc) | is.na(bmi)) %>%
  nrow()
```

## Demographic

```{r}
dat_demographic <- setDT(dat_demo)
dat_demographic <- dat_demographic[dat_demographic$PATID %in% dat_phase4$PATID, ]
dat_demographic <- dat_demographic[, .(PATID, BIRTH_DATE, DEATH_DATE, sex = SEX, race = RACE)]
cohort <- left_join(cohort, dat_demographic, by = "PATID")
cohort[, age := (as.Date(INDEX_DATE) - as.Date(BIRTH_DATE)) / 365]
cohort[, time2death := (as.Date(DEATH_DATE) - as.Date(INDEX_DATE)) / 365]
cohort[, BIRTH_DATE := NULL]
cohort[, DEATH_DATE := NULL]
cohort[, age := as.numeric(age)]
cohort[, time2death := as.numeric(time2death)]
```

## Event

```{r}
#Codeset for outcomes
mi_codes_icd9 <- c("410")
stroke_codes_icd9=c("431", "432", "433", "434")
hf_codes_icd9 <- c("428")
mi_codes_icd10 <- c("I21", "I22", "I23")
stroke_codes_icd10=c("I61", "I62", "I63")
hf_codes_icd10 <- c("I50")
```


```{r}
# use icd code to filter event
dat_cvd_lookup <- dat_diagnosis_cvd1[dat_diagnosis_cvd1$PATID %in% dat_phase4$PATID, ]
dat_event0=dat_cvd_lookup
dat_event0$event=NA
for(nm in c("mi","stroke", "hf")){
  print(nm)
  nm.working1=paste0(nm, "_codes_icd9")
  nm.working2=paste0(nm, "_codes_icd10")
  
  xx=paste(unlist(lapply(c(get(nm.working1),get(nm.working2)), function(x) paste0("^",x))), collapse="|")
  
  ind <- which(grepl(xx,dat_event0$DX))
  dat_event0$event[ind]=nm
}

# join for INDEX DATE
dat_event0 <- left_join(dat_event0, cohort[, .(PATID, INDEX_DATE)], by = 'PATID')

# get diff year
setDT(dat_event0)
dat_event0[, diff_year := as.numeric(as.Date(ADMIT_DATE) - as.Date(INDEX_DATE)) / 365]

# mi
dat_mi=dat_event0[dat_event0$event=="mi",]
dat_mi<- dat_mi[dat_mi[, .I[which.min(diff_year)], by = PATID]$V1]
dat_mi=dat_mi[, .(PATID, diff_year)]
colnames(dat_mi)[2]="time2mi"

# stroke
dat_stroke=dat_event0[dat_event0$event=="stroke",]
dat_stroke<- dat_stroke[dat_stroke[, .I[which.min(diff_year)], by = PATID]$V1]
dat_stroke=dat_stroke[, .(PATID, diff_year)]
colnames(dat_stroke)[2]="time2stroke"

# hf
dat_hf=dat_event0[dat_event0$event=="hf",]
dat_hf<- dat_hf[dat_hf[, .I[which.min(diff_year)], by = PATID]$V1]
dat_hf=dat_hf[, .(PATID, diff_year)]
colnames(dat_hf)[2]="time2hf"

# join all time2 event
cohort <- left_join(cohort, dat_mi, by = "PATID")
cohort <- left_join(cohort, dat_stroke, by = "PATID")
cohort <- left_join(cohort, dat_hf, by = "PATID")

# follow up info
setDT(cohort)
dat_followup_info <- dat_followup_wide[dat_followup_wide$PATID %in% dat_phase4$PATID,]
cohort <- left_join(cohort, dat_followup_info[, .(PATID, end_date)], by = 'PATID')
cohort[, followup_years := as.numeric(as.Date(end_date) - as.Date(INDEX_DATE)) / 365]
cohort[, end_date := NULL]
```


## Save data
```{r}
write.csv(cohort, '../data/raw.csv', row.names = FALSE)
```


#### Data Cleaning

## Oracle Database Connection
```{r warning = F, message = F}
# install RODBC package 
# install.packages ("RODBC", repo="http://archive.linux.duke.edu/cran/") 
# load RODBC package 
library(RODBC)
channel <- odbcConnect("DSR_PROD", believeNRows=FALSE) 
sqlTables(channel, schema = "BI_ACE_VIEWS", tableType = "VIEW", tableName = "PCOR_RSH_%") 
```

## EDA - Missing Values

```{r}
raw = read.csv('../../data/strict_cohort/raw.csv')

# eda for missing values
colSums(is.na(raw))
```



## Binary Indicator and Missing Values

```{r}
# event 10 year
raw$stroke_10y = 1 - is.na(raw$time2stroke)
raw$mi_10y = 1 - is.na(raw$time2mi)
raw$hf_10y = 1 - is.na(raw$time2hf)
raw$death_10y = 1 - is.na(raw$time2death)
raw$cvd_10y = 1 * ((raw$stroke_10y + raw$mi_10y + raw$hf_10y) > 0)


# event 5 year
raw$stroke_5y <- ifelse(!is.na(raw$time2stroke) & raw$time2stroke <= 5, 1, 0)
raw$mi_5y <- ifelse(!is.na(raw$time2mi) & raw$time2mi <= 5, 1, 0)
raw$hf_5y <- ifelse(!is.na(raw$time2hf) & raw$time2hf <= 5, 1, 0)
raw$death_5y <- ifelse(!is.na(raw$time2death) & raw$time2death <= 5, 1, 0)
raw$cvd_5y <- 1 * ((raw$stroke_5y + raw$mi_5y + raw$hf_5y) > 0)

# gender
raw$sex = ifelse(raw$sex == "F", 1, 0)

# fills missing time-to-event values with total follow-up years
raw$time2mi[is.na(raw$time2mi)] = raw$followup_years[is.na(raw$time2mi)]
raw$time2stroke[is.na(raw$time2stroke)] = raw$followup_years[is.na(raw$time2stroke)]
raw$time2hf[is.na(raw$time2hf)] = raw$followup_years[is.na(raw$time2hf)]
raw$time2death[is.na(raw$time2death)] = raw$followup_years[is.na(raw$time2death)]


setnames(raw, old = c("time2mi", "time2stroke", "time2hf", "time2death"),
              new = c("time2mi_10y", "time2stroke_10y", "time2hf_10y", "time2death_10y"))

raw$time2mi_5y     <- pmin(raw$time2mi_10y, 5)
raw$time2stroke_5y <- pmin(raw$time2stroke_10y, 5)
raw$time2hf_5y     <- pmin(raw$time2hf_10y, 5)
raw$time2death_5y  <- pmin(raw$time2death_10y, 5)

# cvd percentage
sum(raw$cvd_10y)/length(raw$cvd_10y)

colSums(is.na(raw))
```

## Calculare eGFR Based on Formula

```{r}
# define function to calculate eGFR
calculate_eGFR <- function(sex, scre, age) {
  if (sex == 1) {  # Female
    if (scre <= 0.7) {
      gfr <- 142 * (scre / 0.7)^-0.241 * 0.9938^age
    } else {
      gfr <- 142 * (scre / 0.7)^-1.2 * 0.9938^age
    }
  } else {  # Male
    if (scre <= 0.9) {
      gfr <- 142 * (scre / 0.9)^-0.302 * 0.9938^age
    } else {
      gfr <- 142 * (scre / 0.9)^-1.2 * 0.9938^age
    }
  }
  return(gfr)
}

# apply function to the df
raw$eGFR <- mapply(calculate_eGFR, raw$sex, raw$scre, raw$age)

# view the updated df with the new eGFR variable
head(raw)
```


## Apply PREVENT Exclusion Criteria

```{r}
# 127839 patients satisfy PREVENT Criteria
raw <- raw %>%
  filter(age >= 30 & age <= 79) %>%
  filter(sbp >= 90 & sbp <= 200) %>%
  filter(tc >= 130 & tc <= 320) %>%
  filter(hdlc >= 20 & hdlc <= 100) %>%
  filter(bmi >= 18.5 & bmi < 40.0)

length(unique(raw$PATID))
```


## SDOH - PCOR_RSH_ADDR_GEO_VW

```{r}
dat_sdoh = NULL
dat_sdoh = sqlQuery(channel, 
paste0("SELECT PATID, ADDRESS_PERIOD_START, ADDRESS_PERIOD_END, ADI_NATRANK, ADI_STATERNK, ADI_VERSION
FROM BI_ACE_VIEWS.PCOR_RSH_ADDR_GEO_VW"))

# filter sdoh data for only final_pat cohort
final_pat = raw$PATID
dat_sdoh=dat_sdoh[which(dat_sdoh$PATID%in%final_pat),]

# save sdoh data
write.csv(dat_sdoh, "../data/dat_sdoh.csv")
```



## Smoker Data - Vital

```{r}
dat_vital=NULL
# loop over all years and concat data
for(my_year in c(14:23)){
print(my_year)
load(file =paste0("P:/Pro00110219 - Predictive modeling using EHR/ch513/PREVENT/Data/dat_vital_cvd",my_year,".RData"))
dat_vital=rbind(dat_vital, dat_vital_cvd)
}

# filter vital data for only final_pat cohort
final_pat = raw$PATID
dat_vital=dat_vital[which(dat_vital$PATID%in%final_pat),]

# save
write.csv(dat_vital, "../data/dat_vital.csv")
```


```{r}
# check missing smoker data in vital
missing = c('NI','UN','OT')
length(dat_vital[dat_vital$SMOKING %in% missing,]$SMOKING)/length(dat_vital$SMOKING)
```


## Smoker Data - ICD10

```{r}
dat_smoke = NULL
final_pat = raw$PATID
# loop over all years
for (my_year in c(14:23)) {
    print(my_year)
    load(file = paste0("P:/Pro00110219 - Predictive modeling using EHR/ch513/PREVENT/Data/dat_diagnosis", my_year, ".RData"))
    # filter by final cohort & ICD code for smoker in diagnosis
    filtered_data <- dat_diagnosis %>% 
      filter(PATID %in% final_pat & grepl("^Z53\\.01|^O99\\.33|^F17|^Z72\\.0", DX)) %>% 
      select(PATID, ADMIT_DATE, DX)
    dat_smoke = rbind(dat_smoke, filtered_data)
}

# save
write.csv(dat_smoke, '../data/dat_smoke.csv')
```

## Diabetes Data - ICD 10

```{r}
dat_db = NULL
final_pat = raw$PATID
# loop over all years
for (my_year in c(14:23)) {
    print(my_year)
    load(file = paste0("P:/Pro00110219 - Predictive modeling using EHR/ch513/PREVENT/Data/dat_diagnosis", my_year, ".RData"))
    # filter data by final cohort & Diabetes ICD code in diagnosis
    filtered_data <- dat_diagnosis %>% 
      filter(PATID %in% final_pat & grepl("^E1[013]", DX)) %>% 
      select(PATID, ADMIT_DATE, DX)
    dat_db = rbind(dat_db, filtered_data)
}

# save
write.csv(dat_db, '../data/dat_db.csv')
```


## Med Data - Statin

```{r}
# define statin RXNORM codes
statin_codes <- c("1790679","1944264","197903", "197904","197905","198211",
                  "2001254","2001262", "2001266","200345","2167557",
                  "2167565","2167569","2167573", "259255","310404","310405",
                  "312961","312962","314231", "359731","359732",
                  "360507","404011", "404013","433849",
                  "476345","476349", "476350","476351",
                  "597967","597971", "597974","597977",
                  "597980","597984", "597987","597990",
                  "597993","617310", "617311","617312","859419",
                  "859424","859747","859751","861643", "861648","861652",
                  "904458","904467", "904475","904481")

# covert to single regular expression pattern
pattern <- paste0("^(", paste(statin_codes, collapse = "|"), ")")

dat_statin = NULL
final_pat = raw$PATID
# loop over all years
for (my_year in c(15:23)) {
    print(my_year)
    load(file = paste0("P:/Pro00110219 - Predictive modeling using EHR/ch513/PREVENT/Data/dat_med", my_year, ".RData"))
    # filter by final cohort and statin EXNORM code in medication
    filtered_data <- dat_med %>% 
      filter(PATID %in% final_pat & grepl(pattern, RXNORM_CUI)) %>%
      select(PATID, RX_ORDER_DATE, RXNORM_CUI)
    dat_statin = rbind(dat_statin, filtered_data)
}

# save
write.csv(dat_statin, '../data/dat_statin.csv')
```

##  Med Data - Anti-Hypertensive

```{r}
# define anti-hypertensive RXNORM code
antihtn_codes <- c("10763","1369","1808","2396","2409","38413","4109","4603","5487","5495","5764","644",
                      "6628","6916","8565","1191185","1297753","1297757", "1495058","1593725","1606347",
                      "1606349","1798281","1923422","1923424","1923426","197379","197380","197381","197382",
                      "197383","198000","198001","198006","198007","198008","198104","198105","198284","198285",
                      "198286","200031","200032","200033","387013","686924","751612","751618","827073",
                      "854901","854905","854908","854916","854919","856422","856429","856448","856457",
                      "856460","856481","856519","856535","856556","856569","856578","856724","856733",
                      "860510","860516","860522","860532","866412","866419","866427","866436","866452",
                      "866461","866472","866479","866482","866491","866511","866514","866924","896758",
                      "896762","896766","904589","998685","998689","17767","233603","33910","4316",
                      "7396","7417","7426","7435","11170","3443","1000001","1091646","1091652","1235144",
                      "1235151","1299859","1299871","1299890","1299896","1299897","1435624",
                      "153822","153823","1600716","1600724","1600728","1656340","1656349","1656354",
                      "197436","197437","197438","197439","197884","197885","197886","197887","198188","198189",
                      "199351","199352","199353","200094","200095","200096","200284","200285","205304","205305",
                      "205326","261962","282755","283316","283317","308962","308963","308964","310140","310792",
                      "310793","310796","310797","310809","311353","311354","312748","312749",
                      "312750","314076","314077","314203","317173","349199","349200","349201",
                      "349353","349373","349401","349405","349483","351292","351293","403853","403854",
                      "403855","477130","485471","577776","578325","578330","636042","636045",
                      "639537","722126","722131","722134","722137","730861","730866","730869","730872",
                      "802749","845488","848131","848135","848140","848145","848151","854925","854984",
                      "854988","857166","857169","857174","857183","857187","858804","858810","858813",
                      "858817","858824","858828","876514","876519","876524","876529","897781","897783",
                      "897844","897853","898342","898346","898350","898353","898356","898359","898362",
                      "898367","898372","898378","898687","898690","898719","898723","979464","979468",
                      "979471","979480","979485","979492","999967","999986","999991","999996","198196",
                      "198197","6876")

# convert to single regular expression pattern
pattern <- paste0("^(", paste(antihtn_codes, collapse = "|"), ")")

dat_antihtn = NULL
final_pat = raw$PATID
# loop through all years
for (my_year in c(15:23)) {
    print(my_year)
    load(file = paste0("P:/Pro00110219 - Predictive modeling using EHR/ch513/PREVENT/Data/dat_med", my_year, ".RData"))
    # filter for final cohort and anti-hypertensive RXNORM code
    filtered_data <- dat_med %>% 
      filter(PATID %in% final_pat & grepl(pattern, RXNORM_CUI)) %>%
      select(PATID, RX_ORDER_DATE, RXNORM_CUI)
    dat_antihtn = rbind(dat_antihtn, filtered_data)
}

# save
write.csv(dat_antihtn, '../data/dat_antihtn.csv')
```


## Insurance Type

```{r}
final_pat <- raw$PATID

# chunk size
chunk_size <- 500

# split final_pat into batches
batches <- split(final_pat, ceiling(seq_along(final_pat) / chunk_size))

# init empty list
dat_ins_list <- list()

# loop through batches
for (i in seq_along(batches)) {
  print(i)
  batch_ids <- batches[[i]]
  
  batch_str <- paste(sprintf("'%s'", batch_ids), collapse = ",")
  
  # sql query for this batch
  query <- paste0("SELECT PATID, ADMIT_DATE, RAW_PAYER_TYPE_PRIMARY FROM BI_ACE_VIEWS.PCOR_RSH_ENC_VW WHERE PATID IN (",batch_str, ")")

  
  # query data and append to list
  dat_ins_list[[i]] <- sqlQuery(channel, query)
}

# combine
dat_ins <- do.call(rbind, dat_ins_list)

# remove missing
dat_ins <- na.omit(dat_ins)

# save insurance data
write.csv(dat_ins, "../data/dat_ins.csv")
```


## Data Integration

For diabetes, statin, and anti-hypertensive, label patient as 1 if there is such a record code within 3 years look back window of the Index Date of that patient.

## Diabetes

```{r}
# load the dataset
dat_db <- fread('../../data/strict_cohort/dat_db.csv')

# use data.table for faster operation
setDT(dat_db)
setDT(raw)

# convert ADMIT_DATE to Date time format
dat_db[, ADMIT_DATE := as.Date(ADMIT_DATE, format = "%Y-%m-%d")]

# ensure INDEX_DATE is in Date time format
raw[, INDEX_DATE := as.Date(INDEX_DATE, format = "%Y-%m-%d")]

# join 'raw' and 'dat_db' on 'PATID'
merged_data <- merge(raw, dat_db, by = "PATID", allow.cartesian = TRUE)

# compare ADMIT_DATE with INDEX_DATE and create the 'diabetes' column
merged_data[, diabetes := as.integer(
  ADMIT_DATE <= as.Date(INDEX_DATE) & 
  ADMIT_DATE >= as.Date(INDEX_DATE) - as.difftime(3 * 365, units = "days")
)]


# summarize by PATID, setting 'diabetes' to 1 if any condition is met
result <- merged_data[, .(diabetes = max(diabetes)), by = .(PATID)]

# join the summarized result back into 'raw'
raw <- merge(raw, result, by = "PATID", all.x = TRUE)

# replace NA values with 0 (for rows with no matching condition)
raw[is.na(diabetes), diabetes := 0]

# Diabetes Distribution: ~11.35% diabetes
table(raw$diabetes)
```



## Medication - Statin

```{r}
# load the dataset
dat_statin = fread('../../data/strict_cohort/dat_statin.csv')

# convert RX_ORDER_DATE to date time format
dat_statin$RX_ORDER_DATE <- as.Date(dat_statin$RX_ORDER_DATE, format = "%Y-%m-%d")

# use data.table for faster operations on large datasets
setDT(dat_statin)
setDT(raw)

# join 'raw' and 'dat_statin' on 'PATID'
merged_data <- merge(raw, dat_statin, by = "PATID", allow.cartesian = TRUE)

# compare dates and create the 'statin' column based on the condition
merged_data[, statin := as.integer(
  RX_ORDER_DATE <= as.Date(INDEX_DATE) & 
  RX_ORDER_DATE >= as.Date(INDEX_DATE) - as.difftime(3 * 365, units = "days")
)]


# summarize the result for each 'PATID' in the 'raw' dataset
result <- merged_data[, .(statin = max(statin)), by = .(PATID)]

# join 'statin' column back into the 'raw' dataframe
raw <- merge(raw, result, by = "PATID", all.x = TRUE)

# replace NA values with 0 in case there were no matching rows
raw$statin[is.na(raw$statin)] <- 0

# Statin Distribution: ~22.48% statin
table(raw$statin)
```


## Medication - Anti-Hypertensive

```{r}
# load dataset
dat_antihtn = fread('../../data/strict_cohort/dat_antihtn.csv')

# convert RX_ORDER_DATE to date format
dat_antihtn$RX_ORDER_DATE <- as.Date(dat_antihtn$RX_ORDER_DATE, format = "%Y-%m-%d")

# use data.table for faster operations on large datasets
setDT(dat_antihtn)
setDT(raw)

# join 'raw' and 'dat_antihtn' on 'PATID'
merged_data <- merge(raw, dat_antihtn, by = "PATID", allow.cartesian = TRUE)

# compare dates and create the 'antihtn' column based on the condition
merged_data[, antihtn := as.integer(
  RX_ORDER_DATE <= as.Date(INDEX_DATE) & 
  RX_ORDER_DATE >= as.Date(INDEX_DATE) - as.difftime(3 * 365, units = "days")
)]

# summarize the result for each 'PATID' in the 'raw' dataset
result <- merged_data[, .(antihtn = max(antihtn)), by = .(PATID)]

# join 'antihtn' column back into the 'raw' dataframe
raw <- merge(raw, result, by = "PATID", all.x = TRUE)

# replace NA values with 0 in case there were no matching rows
raw$antihtn[is.na(raw$antihtn)] <- 0

# Anti-Hypertensive Distribution: ~36.22% antihtn
table(raw$antihtn)
```


## Smoker

Vital Dataset: 

01=Current every day smoker; 02=Current some day smoker;

03=Former smoker; 04=Never smoker;

05=Smoker, current status unknown; 06=Unknown if ever smoked; 07=Heavy tobacco smoker; 08=Light tobacco smoker;
NI=No information; UN=Unknown; OT=Other


Only consider 01, 02, 03, and 04 since they provide clear information on current smoking status

```{r}
dat_vital = fread('../../data/strict_cohort/dat_vital.csv')
dat_smoke = fread('../../data/strict_cohort/dat_smoke.csv')
table(dat_vital$SMOKING)
length(unique(dat_smoke$PATID))
```


```{r}
# convert data to data.table for faster processing
setDT(raw)
setDT(dat_vital)
setDT(dat_smoke)

# convert date columns to time format
dat_smoke[, ADMIT_DATE := as.Date(ADMIT_DATE, format = "%Y-%m-%d")]
dat_vital[, MEASURE_DATE := as.Date(MEASURE_DATE, format = "%Y-%m-%d")]

# step 1: label smoking status(current)
# mark all records in dat_smoke as 1
dat_smoke[, smoker := 1]
# mark records in dat_vital based on SMOKING value
dat_vital[SMOKING %in% c("01", "02"), smoker := 1]
dat_vital[SMOKING %in% c("03", "04"), smoker := 0]

# step 2: stack the data together
dat_smoke_stacked <- dat_smoke[, .(PATID, DATE = ADMIT_DATE, smoker)]
dat_vital_stacked <- dat_vital[, .(PATID, DATE = MEASURE_DATE, smoker)]
# combine both data sources into one
combined_data <- rbind(dat_smoke_stacked, dat_vital_stacked)

# step 3: set key for faster operation: sort by PATID, then DATE
setkey(combined_data, PATID, DATE)
setkey(raw, PATID, INDEX_DATE)
# join combined data with raw on PATID
merged_data <- combined_data[raw, on = .(PATID), allow.cartesian = TRUE]

# step 4: filter for dates before and after the INDEX_DATE
merged_data[, `:=`(diff_before = as.numeric(INDEX_DATE - DATE),  # difference in days before
                   diff_after = as.numeric(DATE - INDEX_DATE))]  # difference in days after

# filter out rows where diff_before > 3 years or diff_after > 3 years
three_years <- 3 * 365
merged_data <- merged_data[diff_before <= three_years & diff_after <= three_years]


# keep records before and after INDEX_DATE
before_data <- merged_data[diff_before >= 0]
after_data <- merged_data[diff_after >= 0]

# step 5: find the closest record before the INDEX_DATE with a non-missing smoker value
before_closest <- before_data[!is.na(smoker), .SD[which.min(diff_before)], by = .(PATID)]

# if no record exists before, take the closest record after the INDEX_DATE with a non-missing smoker value
after_closest <- after_data[!is.na(smoker), .SD[which.min(diff_after)], by = .(PATID)]


# step 6: combine: label smoker based on closest record before index date, if no, use closest after index date
final_smoker <- rbind(before_closest, after_closest)
final_smoker <- final_smoker[, .SD[which.max(diff_before)], by = .(PATID)]

# step 7: merge the final_smoker column back to raw
raw <- merge(raw, final_smoker[, .(PATID, smoker)], by = "PATID", all.x = TRUE)

# before fillna: 11967 current smokers
table(raw$smoker)
sum(is.na(raw$smoker))  # 235

# fill missing smoker values with 0: assume no data -> not current smoker
raw[is.na(smoker), smoker := 0]

# Current Smoker Distribution: 9.36% smoker
table(raw$smoker)
```



## SDOH

- Missing value in ADDRESS_PERIOD_END indicates CURRENT address according to https://sites.duke.edu/crdm/details-of-data-available/pcor_rsh_addr_geo_vw/

- Rows missing ADI_STATERNK also have ADI_NATRANK missing, so delete those rows

- Some ranks are not in number format(not introduced in website above), so delete those rows

- 2019_ADI is the only version, so we don't include in future steps

```{r}
# load sdoh data
dat_sdoh = fread('../../data/strict_cohort/dat_sdoh.csv')

# check unqiue version
unique(dat_sdoh$ADI_VERSION)

# check missing
colSums(is.na(dat_sdoh))

# check if rows missing ADI_STATERNK also have ADI_NATRANK missing
dat_sdoh %>%
  filter(is.na(ADI_STATERNK) & is.na(ADI_NATRANK)) %>%
  nrow()

# filter out missing
dat_sdoh <- dat_sdoh %>%
  filter(!is.na(ADI_STATERNK) & !is.na(ADI_NATRANK))

# convert rank to integer: rank become NaN if not number, then filter out
dat_sdoh <- dat_sdoh %>%
  mutate(
    ADI_NATRANK = as.integer(ADI_NATRANK),
    ADI_STATERNK = as.integer(ADI_STATERNK)
  ) %>%
  filter(!is.na(ADI_NATRANK) & !is.na(ADI_STATERNK))
```


```{r}
# ensure datatype as date
dat_sdoh <- dat_sdoh %>%
  mutate(
    ADDRESS_PERIOD_END = as.Date(ADDRESS_PERIOD_END),
    ADDRESS_PERIOD_START = as.Date(ADDRESS_PERIOD_START)
  )

raw <- raw %>%
  mutate(INDEX_DATE = as.Date(INDEX_DATE))

# step 1: merge raw with dat_sdoh and filter primary matches:
## Current Address: (Address start <= Index date <= address end) OR (Address start <= Index date  AND address end == NULL)
primary_matches <- raw %>%
  left_join(
    dat_sdoh %>%
      select(PATID, ADDRESS_PERIOD_START, ADDRESS_PERIOD_END, ADI_NATRANK, ADI_STATERNK),
    by = "PATID"
  ) %>%
  mutate(
    PRIMARY_MATCH = ADDRESS_PERIOD_START <= INDEX_DATE & 
      (is.na(ADDRESS_PERIOD_END) | INDEX_DATE <= ADDRESS_PERIOD_END)
  ) %>%
  filter(PRIMARY_MATCH) %>%
  group_by(PATID, INDEX_DATE) %>%
  slice_min(ADDRESS_PERIOD_START) %>%
  mutate(DATE_DIFF = 0) %>%
  ungroup()

# step 2: Filter rows without primary matches and find secondary matches
## Not Current: Find closest record(either address start or end) to Index date(Automatically missing for PATID not in dat_sdoh)
secondary_matches <- raw %>%
  anti_join(primary_matches, by = c("PATID", "INDEX_DATE")) %>%
  left_join(
    dat_sdoh %>%
      select(PATID, ADDRESS_PERIOD_START, ADDRESS_PERIOD_END, ADI_NATRANK, ADI_STATERNK),
    by = "PATID") %>%
  mutate(
    DATE_DIFF = pmin(
      abs(as.numeric(INDEX_DATE - ADDRESS_PERIOD_START)),
      abs(as.numeric(INDEX_DATE - ADDRESS_PERIOD_END)),
      na.rm = TRUE
    )) %>%
  group_by(PATID, INDEX_DATE) %>%
  slice_min(DATE_DIFF, with_ties = FALSE) %>%
  ungroup()

# step 3: combine primary and secondary matches
final_merged <- bind_rows(
  primary_matches %>%
    select(-ADDRESS_PERIOD_START, -ADDRESS_PERIOD_END, -PRIMARY_MATCH),
  secondary_matches %>%
    select(-ADDRESS_PERIOD_START, -ADDRESS_PERIOD_END))



# 687 patients missing in SDOH data
raw <- final_merged
colSums(is.na(raw))

hist(raw$DATE_DIFF[!is.na(raw$DATE_DIFF) & raw$DATE_DIFF != 0], breaks = 100,
     main = "Distribution of DATE_DIFF (Excluding NA and 0)", xlab = "DATE_DIFF", ylab = "Frequency")
```



## Insurance

Use the insurance type record with minimal absolute difference in time between Index Date of that patient, leave as missing if no record available for that patient

```{r}
# load ins data
dat_ins <- fread('../../data/strict_cohort/dat_ins.csv', check.names = FALSE)
dat_ins <- dat_ins[, -1, with = FALSE]

setDT(raw)
setDT(dat_ins)

# convert to date
dat_ins[, ADMIT_DATE := as.Date(ADMIT_DATE)]
raw[, INDEX_DATE := as.Date(INDEX_DATE)]


# merge raw and dat_ins on PATID to get all possible combinations
merged_data <- merge(raw, dat_ins, by = "PATID", allow.cartesian = TRUE)

# calculate the absolute difference between ADMIT_DATE and INDEX_DATE
merged_data[, abs_diff := abs(ADMIT_DATE - INDEX_DATE)]

# for each PATID in the merged data, keep only the row with the minimum absolute difference
merged_data <- merged_data[merged_data[, .I[which.min(abs_diff)], by = PATID]$V1]

# add the selected RAW_PAYER_TYPE_PRIMARY back to raw as insurance_type
raw <- merge(
  raw, 
  merged_data[, .(PATID, RAW_PAYER_TYPE_PRIMARY)], 
  by = "PATID", 
  all.x = TRUE
)

# rename column
setnames(raw, "RAW_PAYER_TYPE_PRIMARY", "insurance_type")
```


## Data Transformation


```{r}
# SDOH Non-Missing
raw <- raw[!is.na(ADI_NATRANK) & !is.na(insurance_type)]

raw[, DATE_DIFF := NULL]

# save
write.csv(raw, '../../data/strict_cohort2/cleaned_data.csv', row.names = FALSE)

# load
data = fread('../../data/strict_cohort2/cleaned_data.csv')

# transform
data$time2ascvd_10y <- pmin(data$time2mi_10y, data$time2stroke_10y)
data$ascvd_10y <- 1*((data$stroke_10y+data$mi_10y)>0)
data$time2cvd_10y <- pmin(data$time2mi_10y, data$time2stroke_10y, data$time2hf_10y)

data$time2ascvd_5y <- pmin(data$time2mi_5y, data$time2stroke_5y)
data$ascvd_5y <- 1*((data$stroke_5y+data$mi_5y)>0)
data$time2cvd_5y <- pmin(data$time2mi_5y, data$time2stroke_5y, data$time2hf_5y)

# calculate transformed 'age' column
data$age <- (data$age - 55) / 10

# calculate 'nonhdlc' based on 'tc' and 'hdlc'
data$nonhdlc <- (data$tc - data$hdlc) * 0.02586 - 3.5

# transform 'hdlc'
data$hdlc <- (data$hdlc * 0.02586 - 1.3) / 0.3

# calculate 'sbp1' using the min value of 'sbp' and 110
data$sbp1 <- (pmin(data$sbp, 110) - 110) / 20

# calculate 'sbp2' using the max value of 'sbp' and 110
data$sbp2 <- (pmax(data$sbp, 110) - 130) / 20

# calculate 'bmi1' using the min value of 'bmi' and 30
data$bmi1 <- (pmin(data$bmi, 30) - 25) / 5

# calculate 'bmi2' using the max value of 'bmi' and 30
data$bmi2 <- (pmax(data$bmi, 30) - 30) / 5

# calculate 'egfr1' using the min value of 'egfr' and 60
data$egfr1 <- (pmin(data$eGFR, 60) - 60) / -15

# calculate 'egfr2' using the max value of 'egfr' and 60
data$egfr2 <- (pmax(data$eGFR, 60) - 90) / -15

# calculate 'sbp2treat' as the product of 'sbp2' and 'antihtn'
data$sbp2treat <- data$sbp2 * data$antihtn

# calculate 'nonhdlctreat' as the product of 'nonhdlc' and 'statin'
data$nonhdlctreat <- data$nonhdlc * data$statin

# calculate 'agenonhdlc' as the product of 'age' and 'nonhdlc'
data$agenonhdlc <- data$age * data$nonhdlc

# calculate 'agehdlc' as the product of 'age' and 'hdlc'
data$agehdlc <- data$age * data$hdlc

# calculate 'agesbp2' as the product of 'age' and 'sbp2'
data$agesbp2 <- data$age * data$sbp2

# calculate 'agediabetes' as the product of 'age' and 'diabetes'
data$agediabetes <- data$age * data$diabetes

# calculate 'agecursmk' as the product of 'age' and 'cursmk'
data$agecursmk <- data$age * data$smoker

# calculate 'agebmi2' as the product of 'age' and 'bmi2'
data$agebmi2 <- data$age * data$bmi2

# calculate 'ageegfr1' as the product of 'age' and 'egfr1'
data$ageegfr1 <- data$age * data$egfr1

# set 'cons' column to 1 for all rows: for intercept
data$cons <- 1

# insurance
data$insurance_group <- with(data, ifelse(insurance_type %in% c("COMMERCIAL", "MANAGED CARE", "NC BLUE CROSS", "OOS BLUE CROSS"),
                                          "Commercial",
                                   ifelse(insurance_type %in% c("MEDICARE", "MEDICARE ADVANTAGE"),
                                          "Medicare",
                                   ifelse(insurance_type %in% c("MEDICAID PENDING", "NC MEDICAID", "OOS MEDICAID"),
                                          "Medicaid",
                                   ifelse(insurance_type == "SELF-PAY",
                                          "No Insurance",
                                          "Other Insurance")))))

# split SDOH into quartiles
data <- data %>%
  mutate(
    ADI_NATRANK_BIN = cut(
      ADI_NATRANK,
      breaks = seq(0, 100, by = 25),
      labels = 1:4,
      include.lowest = TRUE))

# save
write.csv(data, '../../data/strict_cohort2/transformed_data.csv', row.names = FALSE)
```




## PREVENT


```{r}
# 5 year coefficients
coefficients_5y <- list(
  female_cvd = matrix(c(
    0.7939329, 0.0305239, -0.1606857, -0.2394003, 0.360078, 0.8667604,
    0.5360739, 0.6045917, 0.0433769, 0.3151672, -0.1477655, -0.0663612,
    0.1197879, -0.0819715, 0.0306769, -0.0946348, -0.27057, -0.078715,
    -0.1637806, -4.165859
  ), nrow = 20, ncol = 1),
  male_cvd = matrix(c(
    0.7688528, 0.0736174, -0.0954431, -0.4347345, 0.3362658, 0.7692857,
    0.4386871, 0.5378979, 0.0164827, 0.288879, -0.1337349, -0.0475924,
    0.150273, -0.0517874, 0.0191169, -0.1049477, -0.2251948, -0.0895067,
    -0.1543702, -3.889299
  ), nrow = 20, ncol = 1),
  female_ascvd = matrix(c(
    0.7198830247, 0.1176967025, -0.1511850059, -0.0835357979, 0.3592852056,
    0.8348584771, 0.4831078053, 0.4864619076, 0.039777901, 0.2265308946,
    -0.0592373982, -0.0395761989, 0.0844423026, -0.0567838997,
    0.0325691998, -0.1035984978, -0.241754204, -0.0791141987,
    -0.167149201, -4.691118
  ), nrow = 20, ncol = 1),
  male_ascvd = matrix(c(
    0.7099847198, 0.1658663005, -0.1144284979, -0.2837212086, 0.3239977062,
    0.7189596891, 0.3956972957, 0.369007498, 0.0203619003, 0.2036522031,
    -0.0865581036, -0.0322915986, 0.1145630032, -0.0300005004,
    0.0232746992, -0.0927024037, -0.2018525004, -0.0970527008,
    -0.1217081025, -4.371798
  ), nrow = 20, ncol = 1),
  female_hf = matrix(c(
    0.8998234868, -0.4559771121, 0.3576504886, 1.038346052, 0.5839160085,
    -0.0072293999, 0.2997705936, 0.7451637983, 0.0557086989, 0.3534441888,
    -0.0981511027, -0.0946663022, -0.3581041098, -0.1159453019,
    -0.003878, -0.1884288937, -5.23739
  ), nrow = 17, ncol = 1),
  male_hf = matrix(c(
    0.8972641826, -0.6811466217, 0.3634460866, 0.9237759709, 0.5023735762,
    -0.0485840999, 0.3726929128, 0.6926916838, 0.0251826998, 0.2980921865,
    -0.0497731008, -0.1289200932, -0.3040924072, -0.140168801,
    0.0068126, -0.179777801, -4.873372
  ), nrow = 17, ncol = 1)
)

# 10 year coefficients

coefficients_10y <- list(
  female_cvd = matrix(c(
    0.7939329, 0.0305239, -0.1606857, -0.2394003, 0.360078, 0.8667604,
    0.5360739, 0.6045917, 0.0433769, 0.3151672, -0.1477655, -0.0663612,
    0.1197879, -0.0819715, 0.0306769, -0.0946348, -0.27057, -0.078715,
    -0.1637806, -3.307728
  ), nrow = 20, ncol = 1),
  male_cvd = matrix(c(
    0.7688528, 0.0736174, -0.0954431, -0.4347345, 0.3362658, 0.7692857,
    0.4386871, 0.5378979, 0.0164827, 0.288879, -0.1337349, -0.0475924,
    0.150273, -0.0517874, 0.0191169, -0.1049477, -0.2251948, -0.0895067,
    -0.1543702, -3.031168
  ), nrow = 20, ncol = 1),
  female_ascvd = matrix(c(
    0.7198830247, 0.1176967025, -0.1511850059, -0.0835357979, 0.3592852056,
    0.8348584771, 0.4831078053, 0.4864619076, 0.039777901, 0.2265308946,
    -0.0592373982, -0.0395761989, 0.0844423026, -0.0567838997,
    0.0325691998, -0.1035984978, -0.241754204, -0.0791141987,
    -0.167149201, -3.819974899
  ), nrow = 20, ncol = 1),
  male_ascvd = matrix(c(
    0.7099847198, 0.1658663005, -0.1144284979, -0.2837212086, 0.3239977062,
    0.7189596891, 0.3956972957, 0.369007498, 0.0203619003, 0.2036522031,
    -0.0865581036, -0.0322915986, 0.1145630032, -0.0300005004,
    0.0232746992, -0.0927024037, -0.2018525004, -0.0970527008,
    -0.1217081025, -3.500654936
  ), nrow = 20, ncol = 1),
  female_hf = matrix(c(
    0.8998234868, -0.4559771121, 0.3576504886, 1.038346052, 0.5839160085,
    -0.0072293999, 0.2997705936, 0.7451637983, 0.0557086989, 0.3534441888,
    -0.0981511027, -0.0946663022, -0.3581041098, -0.1159453019,
    -0.003878, -0.1884288937, -4.310409069
  ), nrow = 17, ncol = 1),
  male_hf = matrix(c(
    0.8972641826, -0.6811466217, 0.3634460866, 0.9237759709, 0.5023735762,
    -0.0485840999, 0.3726929128, 0.6926916838, 0.0251826998, 0.2980921865,
    -0.0497731008, -0.1289200932, -0.3040924072, -0.140168801,
    0.0068126, -0.179777801, -3.946391106
  ), nrow = 17, ncol = 1)
)

# predictors for each outcome
outcomes <- list(
  cvd = list(
    var = "cvd",
    covars = c(
      "age", "nonhdlc", "hdlc", "sbp1", "sbp2", "diabetes",
      "smoker", "egfr1", "egfr2", "antihtn", "statin",
      "sbp2treat", "nonhdlctreat", "agenonhdlc", "agehdlc",
      "agesbp2", "agediabetes", "agecursmk", "ageegfr1", "cons"
    )
  ),
  ascvd = list(
    var = "ascvd",
    covars = c(
      "age", "nonhdlc", "hdlc", "sbp1", "sbp2", "diabetes",
      "smoker", "egfr1", "egfr2", "antihtn", "statin",
      "sbp2treat", "nonhdlctreat", "agenonhdlc", "agehdlc",
      "agesbp2", "agediabetes", "agecursmk", "ageegfr1", "cons"
    )
  ),
  hf = list(
    var = "hf",
    covars = c(
      "age", "sbp1", "sbp2", "diabetes", "smoker", "bmi1", "bmi2",
      "egfr1", "egfr2", "antihtn", "sbp2treat", "agesbp2",
      "agediabetes", "agecursmk", "agebmi2", "ageegfr1", "cons"
    )
  )
)
```


```{r}
append_predicted_risks <- function(raw, outcomes, coefficients, prefix = "") {
  for (outcome_name in names(outcomes)) {
    new_colname <- paste0(prefix, "_prevent_", outcome_name)
    raw[, (new_colname) := NA_real_]
  }
  
  for (outcome_name in names(outcomes)) {
    covars <- outcomes[[outcome_name]]$covars
    
    for (gender_val in c(0, 1)) {
      gender_str <- ifelse(gender_val == 1, "female", "male")
      coef_key <- paste0(gender_str, "_", outcome_name)
      coef_vec <- coefficients[[coef_key]]
      
      # subset for gender
      subset_idx <- raw$sex == gender_val
      data_subset <- raw[subset_idx]
      X <- as.matrix(data_subset[, ..covars])
      
      # compute predicted probabilities
      logit <- X %*% coef_vec
      prob <- as.numeric(exp(logit) / (1 + exp(logit)))
      
      # assign
      new_colname <- paste0(prefix, "_prevent_", outcome_name)
      raw[subset_idx, (new_colname) := prob]
    }
  }
  
  return(raw)
}


# apply
data <- fread('../../data/strict_cohort2/transformed_data.csv')
data <- append_predicted_risks(data, outcomes, coefficients_5y, prefix = '5y')
data <- append_predicted_risks(data, outcomes, coefficients_10y, prefix = '10y')
```




## Local Models

## Factorize Categorical Variable

```{r}
data$diabetes <- as.factor(data$diabetes)
data$smoker <- as.factor(data$smoker)
data$antihtn <- as.factor(data$antihtn)
data$statin <- as.factor(data$statin)
data$race <- as.factor(data$race)
data$ADI_NATRANK_BIN <- as.factor(data$ADI_NATRANK_BIN)
# set "White" as the reference level due to large sample size
data$race <- relevel(data$race, ref = "White")
```


```{r}
# define function to perform cross-validation for Cox models and get risk scores and probabilities
perform_cv_cox <- function(data, event_var, formula, folds = 5, year = NA) {
  # set seed for reproducibility
  set.seed(123)
  
  # create stratified folds on targeted event
  folds_list <- createFolds(data[[event_var]], k = folds, list = TRUE)
  
  # initialize risk score vector
  risk_scores <- numeric(nrow(data))
  
  # loop over each fold
  for (i in seq_along(folds_list)) {
    # split data into training and testing sets
    test_index <- folds_list[[i]]
    train_data <- data[-test_index, ]
    test_data <- data[test_index, ]
    
    # fit Cox model
    cox_model <- coxph(as.formula(formula), data = train_data)
    cox_lp <- predict(cox_model, newdata = test_data, type = "lp")
    baseline_surv <- survfit(cox_model)
    ten_year <- summary(baseline_surv, times = year, extend = TRUE)$surv
    
    # get risk probabilities on test data
    risk_scores[test_index] <- 1 - ten_year ^ exp(cox_lp)
  }
  
  return(risk_scores)
}
```


## Local Model


```{r}
local_formulas <- list(
  cvd = "Surv(time2cvd_10y, cvd_10y) ~ age + nonhdlc + hdlc + sbp1 + sbp2 +
                           diabetes + smoker + egfr1 + egfr2 + antihtn + statin + sbp2treat +
                           nonhdlctreat + agenonhdlc + agehdlc + agesbp2 + agediabetes +
                           agecursmk + ageegfr1",
  ascvd = "Surv(time2ascvd_10y, ascvd_10y) ~ age + nonhdlc + hdlc + sbp1 + sbp2 +
                             diabetes + smoker + egfr1 + egfr2 + antihtn + statin + sbp2treat +
                             nonhdlctreat + agenonhdlc + agehdlc + agesbp2 + agediabetes +
                             agecursmk + ageegfr1",
  hf = "Surv(time2hf_10y, hf_10y) ~ age + sbp1 + sbp2 +
                         diabetes + smoker + bmi1 + bmi2 + egfr1 + egfr2 + antihtn + sbp2treat +
                         agesbp2 + agediabetes + agecursmk + agebmi2 + ageegfr1")


outcome_list <- names(local_formulas)

# 10 year
prefix <- "10y_local_"
for (outcome_name in outcome_list) {
  for (gender_val in c(0, 1)) {
    # filter gender
    gender_indices <- which(data$sex == gender_val)
    gender_data <- data[gender_indices, ]

    # outcome-specific variables
    event_var <- paste0(outcome_name, "_10y")
    formula_str <- local_formulas[[outcome_name]]
    
    # run cv
    risk <- perform_cv_cox(
      data = gender_data,
      event_var = event_var,
      formula = formula_str,
      folds = 5,
      year = 10)

    # Create output column name and assign to correct rows
    colname <- paste0(prefix, outcome_name)
    data[gender_indices, (colname) := risk]
  }
}

# 5 year
prefix <- "5y_local_"
for (outcome_name in outcome_list) {
  for (gender_val in c(0, 1)) {
    # filter gender
    gender_indices <- which(data$sex == gender_val)
    gender_data <- data[gender_indices, ]

    # outcome-specific variables
    event_var <- paste0(outcome_name, "_10y")
    formula_str <- local_formulas[[outcome_name]]
    
    # run cv
    risk <- perform_cv_cox(
      data = gender_data,
      event_var = event_var,
      formula = formula_str,
      folds = 5,
      year = 5)

    # Create output column name and assign to correct rows
    colname <- paste0(prefix, outcome_name)
    data[gender_indices, (colname) := risk]
  }
}
```



## Local Race Model


```{r}
race_formulas <- list(
  cvd = "Surv(time2cvd_10y, cvd_10y) ~ age + nonhdlc + hdlc + sbp1 + sbp2 +
                           diabetes + smoker + egfr1 + egfr2 + antihtn + statin + sbp2treat +
                           nonhdlctreat + agenonhdlc + agehdlc + agesbp2 + agediabetes +
                           agecursmk + ageegfr1 + race",
  ascvd = "Surv(time2ascvd_10y, ascvd_10y) ~ age + nonhdlc + hdlc + sbp1 + sbp2 +
                             diabetes + smoker + egfr1 + egfr2 + antihtn + statin + sbp2treat +
                             nonhdlctreat + agenonhdlc + agehdlc + agesbp2 + agediabetes +
                             agecursmk + ageegfr1 + race",
  hf = "Surv(time2hf_10y, hf_10y) ~ age + sbp1 + sbp2 +
                         diabetes + smoker + bmi1 + bmi2 + egfr1 + egfr2 + antihtn + sbp2treat +
                         agesbp2 + agediabetes + agecursmk + agebmi2 + ageegfr1 + race")


outcome_list <- names(race_formulas)

# 10 year
prefix <- "10y_race_"
for (outcome_name in outcome_list) {
  for (gender_val in c(0, 1)) {
    # filter gender
    gender_indices <- which(data$sex == gender_val)
    gender_data <- data[gender_indices, ]

    # outcome-specific variables
    event_var <- paste0(outcome_name, "_10y")
    formula_str <- race_formulas[[outcome_name]]
    
    # run cv
    risk <- perform_cv_cox(
      data = gender_data,
      event_var = event_var,
      formula = formula_str,
      folds = 5,
      year = 10)

    # Create output column name and assign to correct rows
    colname <- paste0(prefix, outcome_name)
    data[gender_indices, (colname) := risk]
  }
}

# 5 year
prefix <- "5y_race_"
for (outcome_name in outcome_list) {
  for (gender_val in c(0, 1)) {
    # filter gender
    gender_indices <- which(data$sex == gender_val)
    gender_data <- data[gender_indices, ]

    # outcome-specific variables
    event_var <- paste0(outcome_name, "_10y")
    formula_str <- race_formulas[[outcome_name]]
    
    # run cv
    risk <- perform_cv_cox(
      data = gender_data,
      event_var = event_var,
      formula = formula_str,
      folds = 5,
      year = 5)

    # Create output column name and assign to correct rows
    colname <- paste0(prefix, outcome_name)
    data[gender_indices, (colname) := risk]
  }
}
```



## Local Race and SDOH Model


```{r}
sdoh_formulas <- list(
  cvd = "Surv(time2cvd_10y, cvd_10y) ~ age + nonhdlc + hdlc + sbp1 + sbp2 +
                           diabetes + smoker + egfr1 + egfr2 + antihtn + statin + sbp2treat +
                           nonhdlctreat + agenonhdlc + agehdlc + agesbp2 + agediabetes +
                           agecursmk + ageegfr1 + race + ADI_NATRANK_BIN",
  ascvd = "Surv(time2ascvd_10y, ascvd_10y) ~ age + nonhdlc + hdlc + sbp1 + sbp2 +
                             diabetes + smoker + egfr1 + egfr2 + antihtn + statin + sbp2treat +
                             nonhdlctreat + agenonhdlc + agehdlc + agesbp2 + agediabetes +
                             agecursmk + ageegfr1 + race + ADI_NATRANK_BIN",
  hf = "Surv(time2hf_10y, hf_10y) ~ age + sbp1 + sbp2 +
                         diabetes + smoker + bmi1 + bmi2 + egfr1 + egfr2 + antihtn + sbp2treat +
                         agesbp2 + agediabetes + agecursmk + agebmi2 + ageegfr1 + race + ADI_NATRANK_BIN")


outcome_list <- names(sdoh_formulas)

# 10 year
prefix <- "10y_sdoh_"
for (outcome_name in outcome_list) {
  for (gender_val in c(0, 1)) {
    # filter gender
    gender_indices <- which(data$sex == gender_val)
    gender_data <- data[gender_indices, ]

    # outcome-specific variables
    event_var <- paste0(outcome_name, "_10y")
    formula_str <- sdoh_formulas[[outcome_name]]
    
    # run cv
    risk <- perform_cv_cox(
      data = gender_data,
      event_var = event_var,
      formula = formula_str,
      folds = 5,
      year = 10)

    # Create output column name and assign to correct rows
    colname <- paste0(prefix, outcome_name)
    data[gender_indices, (colname) := risk]
  }
}

# 5 year
prefix <- "5y_sdoh_"
for (outcome_name in outcome_list) {
  for (gender_val in c(0, 1)) {
    # filter gender
    gender_indices <- which(data$sex == gender_val)
    gender_data <- data[gender_indices, ]

    # outcome-specific variables
    event_var <- paste0(outcome_name, "_10y")
    formula_str <- sdoh_formulas[[outcome_name]]
    
    # run cv
    risk <- perform_cv_cox(
      data = gender_data,
      event_var = event_var,
      formula = formula_str,
      folds = 5,
      year = 5)

    # Create output column name and assign to correct rows
    colname <- paste0(prefix, outcome_name)
    data[gender_indices, (colname) := risk]
  }
}
```






## Save

```{r}
# save
write.csv(data, '../../data/strict_cohort2/all_data.csv', row.names = FALSE)
```



